---
title: "Exploring Mushroom Characteristics: Patterns Between Edible and Poisonous Mushrooms"
author:
  - "Anchita Mitra"
  - "Fatmah Almeer"
  - "Heet Jatin Patel"
format: pdf
editor: visual
---

## Introduction:

Apart from their importance in ecosystems, culinary dishes, and biodiversity, mushrooms pose a serious threat to public health in cases where they are misidentified and consumed. Given that there are many different types of mushrooms that look alike, finding characteristics that can differentiate between edible and poisonous mushrooms is both fascinating and relevant. Identifying the differences may allow us to learn how observable attributes can be linked to classification results.

In the current project, our goal is to use the method of exploratory data analysis (EDA) to explore how different characteristics of mushrooms are related to whether or not they are poisonous. As opposed to developing prediction models, our main objective is to find patterns and features that show high predictive potential and compare the distribution of feature values.

## Data Provenance:

The data set employed in this study is the Mushroom Data Set, which is also referred to as the Agaricus-Lepiota mushroom data set. This data set consists of observations about mushrooms based on information about actual species belonging to the Agaricus and Lepiota families. The data set has a number of features, including those regarding the physical characteristics of the mushrooms like cap shape, cap color, odor, gill type, bruising, habitats, and spore print colors.

This data set is classified according to whether the mushrooms are either edible or poisonous. The data were obtained from descriptions found in field guides, and it has been frequently used in classification and pattern recognition research. This data set is especially suitable for use in the current study since it has numerous categorical features.

There are over 8,000 observations in this data set and over 20 attributes, providing enough data for exploratory purposes.

## FAIR and CARE Principles:

### FAIR Principles:

This dataset satisfies several FAIR principles:

-   **Findable:** The mushroom dataset is publicly available and well known in educational and research settings.

-   **Accessible:** The data can be downloaded in standard text format and easily imported into R.

-   **Interoperable:** Because the data are structured in rows and columns, they can be used across multiple software platforms such as R, Python, and spreadsheet tools.

-   **Reusable:** The variables are documented through a Quarto file, allowing future users to understand and reuse the dataset effectively.

### CARE Principles:

The CARE principles emphasize the following: Collective Benefit, Authority to Control, Responsibility, and Ethics. In this case, since the dataset is devoid of any personal and community-held human data, the CARE principles are relevant only to an extent in contrast to social and indigenous datasets. However, in our research, we embrace the spirit of CARE principles through responsible use of the dataset and proper reporting of findings.

## Data Preparation:

The original data file stored each categorical value as a one-character code. We renamed the columns and decoded the attributes and their values using the dataset documentation. After importing the data, we checked the number of rows and columns against the codebook to confirm that the class labels were decoded into the right categories. For the missing data, we switched the values from `?` to `NULL` using `NA_character_` as an indication of a missing value.

-   The cleaned dataset contains 8,124 mushroom cases and 23 variables.

```{r}
#| label: load-and-clean-data
#| echo: false
#| message: false
#| warning: false

# Primary Author: Fatmah
# Reviewer: Heet
# PCIP: Plan to clean the raw local data and improve category names for readability.

library(dplyr)

raw_mushrooms <- read.table("data/agaricus-lepiota.data", header = FALSE, sep = ",")

mushrooms_clean <- raw_mushrooms |>
  rename(
    classes = V1,
    `cap-shape` = V2,
    `cap-surface` = V3,
    `cap-color` = V4,
    `bruises?` = V5,
    odor = V6,
    `gill-attachment` = V7,
    `gill-spacing` = V8,
    `gill-size` = V9,
    `gill-color` = V10,
    `stalk-shape` = V11,
    `stalk-root` = V12,
    `stalk-surface-above-ring` = V13,
    `stalk-surface-below-ring` = V14,
    `stalk-color-above-ring` = V15,
    `stalk-color-below-ring` = V16,
    `veil-type` = V17,
    `veil-color` = V18,
    `ring-number` = V19,
    `ring-type` = V20,
    `spore-print-color` = V21,
    population = V22,
    habitat = V23
  ) |>
  mutate(
    classes = case_when(
      classes == "e" ~ "edible",
      classes == "p" ~ "poisonous",
      classes == "?" ~ NA_character_,
      TRUE ~ classes
    ),
    `cap-shape` = case_when(
      `cap-shape` == "b" ~ "bell",
      `cap-shape` == "c" ~ "conical",
      `cap-shape` == "x" ~ "convex",
      `cap-shape` == "f" ~ "flat",
      `cap-shape` == "k" ~ "knobbed",
      `cap-shape` == "s" ~ "sunken",
      `cap-shape` == "?" ~ NA_character_,
      TRUE ~ `cap-shape`
    ),
    `cap-surface` = case_when(
      `cap-surface` == "f" ~ "fibrous",
      `cap-surface` == "g" ~ "grooves",
      `cap-surface` == "y" ~ "scaly",
      `cap-surface` == "s" ~ "smooth",
      `cap-surface` == "?" ~ NA_character_,
      TRUE ~ `cap-surface`
    ),
    `cap-color` = case_when(
      `cap-color` == "n" ~ "brown",
      `cap-color` == "b" ~ "buff",
      `cap-color` == "c" ~ "cinnamon",
      `cap-color` == "g" ~ "gray",
      `cap-color` == "r" ~ "green",
      `cap-color` == "p" ~ "pink",
      `cap-color` == "u" ~ "purple",
      `cap-color` == "e" ~ "red",
      `cap-color` == "w" ~ "white",
      `cap-color` == "y" ~ "yellow",
      `cap-color` == "?" ~ NA_character_,
      TRUE ~ `cap-color`
    ),
    `bruises?` = case_when(
      `bruises?` == "t" ~ "bruises",
      `bruises?` == "f" ~ "no bruises",
      `bruises?` == "?" ~ NA_character_,
      TRUE ~ `bruises?`
    ),
    odor = case_when(
      odor == "a" ~ "almond",
      odor == "l" ~ "anise",
      odor == "c" ~ "creosote",
      odor == "y" ~ "fishy",
      odor == "f" ~ "foul",
      odor == "m" ~ "musty",
      odor == "n" ~ "none",
      odor == "p" ~ "pungent",
      odor == "s" ~ "spicy",
      odor == "?" ~ NA_character_,
      TRUE ~ odor
    ),
    `gill-attachment` = case_when(
      `gill-attachment` == "a" ~ "attached",
      `gill-attachment` == "d" ~ "descending",
      `gill-attachment` == "f" ~ "free",
      `gill-attachment` == "n" ~ "notched",
      `gill-attachment` == "?" ~ NA_character_,
      TRUE ~ `gill-attachment`
    ),
    `gill-spacing` = case_when(
      `gill-spacing` == "c" ~ "close",
      `gill-spacing` == "w" ~ "crowded",
      `gill-spacing` == "d" ~ "distant",
      `gill-spacing` == "?" ~ NA_character_,
      TRUE ~ `gill-spacing`
    ),
    `gill-size` = case_when(
      `gill-size` == "b" ~ "broad",
      `gill-size` == "n" ~ "narrow",
      `gill-size` == "?" ~ NA_character_,
      TRUE ~ `gill-size`
    ),
    `gill-color` = case_when(
      `gill-color` == "k" ~ "black",
      `gill-color` == "n" ~ "brown",
      `gill-color` == "b" ~ "buff",
      `gill-color` == "h" ~ "chocolate",
      `gill-color` == "g" ~ "gray",
      `gill-color` == "r" ~ "green",
      `gill-color` == "o" ~ "orange",
      `gill-color` == "p" ~ "pink",
      `gill-color` == "u" ~ "purple",
      `gill-color` == "e" ~ "red",
      `gill-color` == "w" ~ "white",
      `gill-color` == "y" ~ "yellow",
      `gill-color` == "?" ~ NA_character_,
      TRUE ~ `gill-color`
    ),
    `stalk-shape` = case_when(
      `stalk-shape` == "e" ~ "enlarging",
      `stalk-shape` == "t" ~ "tapering",
      `stalk-shape` == "?" ~ NA_character_,
      TRUE ~ `stalk-shape`
    ),
    `stalk-root` = case_when(
      `stalk-root` == "b" ~ "bulbous",
      `stalk-root` == "c" ~ "club",
      `stalk-root` == "u" ~ "cup",
      `stalk-root` == "e" ~ "equal",
      `stalk-root` == "z" ~ "rhizomorphs",
      `stalk-root` == "r" ~ "rooted",
      `stalk-root` == "?" ~ NA_character_,
      TRUE ~ `stalk-root`
    ),
    `stalk-surface-above-ring` = case_when(
      `stalk-surface-above-ring` == "f" ~ "fibrous",
      `stalk-surface-above-ring` == "y" ~ "scaly",
      `stalk-surface-above-ring` == "k" ~ "silky",
      `stalk-surface-above-ring` == "s" ~ "smooth",
      `stalk-surface-above-ring` == "?" ~ NA_character_,
      TRUE ~ `stalk-surface-above-ring`
    ),
    `stalk-surface-below-ring` = case_when(
      `stalk-surface-below-ring` == "f" ~ "fibrous",
      `stalk-surface-below-ring` == "y" ~ "scaly",
      `stalk-surface-below-ring` == "k" ~ "silky",
      `stalk-surface-below-ring` == "s" ~ "smooth",
      `stalk-surface-below-ring` == "?" ~ NA_character_,
      TRUE ~ `stalk-surface-below-ring`
    ),
    `stalk-color-above-ring` = case_when(
      `stalk-color-above-ring` == "n" ~ "brown",
      `stalk-color-above-ring` == "b" ~ "buff",
      `stalk-color-above-ring` == "c" ~ "cinnamon",
      `stalk-color-above-ring` == "g" ~ "gray",
      `stalk-color-above-ring` == "o" ~ "orange",
      `stalk-color-above-ring` == "p" ~ "pink",
      `stalk-color-above-ring` == "e" ~ "red",
      `stalk-color-above-ring` == "w" ~ "white",
      `stalk-color-above-ring` == "y" ~ "yellow",
      `stalk-color-above-ring` == "?" ~ NA_character_,
      TRUE ~ `stalk-color-above-ring`
    ),
    `stalk-color-below-ring` = case_when(
      `stalk-color-below-ring` == "n" ~ "brown",
      `stalk-color-below-ring` == "b" ~ "buff",
      `stalk-color-below-ring` == "c" ~ "cinnamon",
      `stalk-color-below-ring` == "g" ~ "gray",
      `stalk-color-below-ring` == "o" ~ "orange",
      `stalk-color-below-ring` == "p" ~ "pink",
      `stalk-color-below-ring` == "e" ~ "red",
      `stalk-color-below-ring` == "w" ~ "white",
      `stalk-color-below-ring` == "y" ~ "yellow",
      `stalk-color-below-ring` == "?" ~ NA_character_,
      TRUE ~ `stalk-color-below-ring`
    ),
    `veil-type` = case_when(
      `veil-type` == "p" ~ "partial",
      `veil-type` == "u" ~ "universal",
      `veil-type` == "?" ~ NA_character_,
      TRUE ~ `veil-type`
    ),
    `veil-color` = case_when(
      `veil-color` == "n" ~ "brown",
      `veil-color` == "o" ~ "orange",
      `veil-color` == "w" ~ "white",
      `veil-color` == "y" ~ "yellow",
      `veil-color` == "?" ~ NA_character_,
      TRUE ~ `veil-color`
    ),
    `ring-number` = case_when(
      `ring-number` == "n" ~ "none",
      `ring-number` == "o" ~ "one",
      `ring-number` == "t" ~ "two",
      `ring-number` == "?" ~ NA_character_,
      TRUE ~ `ring-number`
    ),
    `ring-type` = case_when(
      `ring-type` == "c" ~ "cobwebby",
      `ring-type` == "e" ~ "evanescent",
      `ring-type` == "f" ~ "flaring",
      `ring-type` == "l" ~ "large",
      `ring-type` == "n" ~ "none",
      `ring-type` == "p" ~ "pendant",
      `ring-type` == "s" ~ "sheathing",
      `ring-type` == "z" ~ "zone",
      `ring-type` == "?" ~ NA_character_,
      TRUE ~ `ring-type`
    ),
    `spore-print-color` = case_when(
      `spore-print-color` == "k" ~ "black",
      `spore-print-color` == "n" ~ "brown",
      `spore-print-color` == "b" ~ "buff",
      `spore-print-color` == "h" ~ "chocolate",
      `spore-print-color` == "r" ~ "green",
      `spore-print-color` == "o" ~ "orange",
      `spore-print-color` == "u" ~ "purple",
      `spore-print-color` == "w" ~ "white",
      `spore-print-color` == "y" ~ "yellow",
      `spore-print-color` == "?" ~ NA_character_,
      TRUE ~ `spore-print-color`
    ),
    population = case_when(
      population == "a" ~ "abundant",
      population == "c" ~ "clustered",
      population == "n" ~ "numerous",
      population == "s" ~ "scattered",
      population == "v" ~ "several",
      population == "y" ~ "solitary",
      population == "?" ~ NA_character_,
      TRUE ~ population
    ),
    habitat = case_when(
      habitat == "g" ~ "grasses",
      habitat == "l" ~ "leaves",
      habitat == "m" ~ "meadows",
      habitat == "p" ~ "paths",
      habitat == "u" ~ "urban",
      habitat == "w" ~ "waste",
      habitat == "d" ~ "woods",
      habitat == "?" ~ NA_character_,
      TRUE ~ habitat
    )
  )
```

## Interpretation of Key Findings:

### Class Distribution:

As we see in @tbl-class. The dataset contains a relatively balanced distribution of edible and poisonous mushrooms (51.8% edible and 48.2% poisonous) . This is useful for exploratory analysis because comparisons between classes are not heavily distorted by extreme imbalance.

```{r}
#| label: tbl-class
#| tbl-cap: "Distribution of edible and poisonous mushrooms in the dataset."
#| echo: false
#| message: false
#| warning: false

# Primary Author: Fatmah
# Reviewer: Anchita
# PCIP: Plan to create a summary of the mushroom classes, with counts and percentages.

library(dplyr)
library(knitr)

class_table <- mushrooms_clean |>
  count(classes, name = "count") |>
  mutate(percentage = paste(round(count / sum(count) * 100, 1), "%"))

class_table |>
  rename(
    Classification = classes,
    Count = count,
    Percentage = percentage
  ) |>
  kable()
```

### Odor and Toxicity:

Odor appears to be one of the strongest indicators of mushroom classification. Mushrooms with foul, pungent, or fishy odors are much more likely to be poisonous, while mushrooms with almond, anise, or no odor are more commonly edible. This suggests odor may be one of the most informative observable characteristics in the dataset. Looking at @fig-odor, we can see the proportion of edible and poisonous mushrooms within each odor category.

```{r}
#| label: fig-odor
#| fig-cap: "Mushroom edibility by odor."
#| fig-alt: "A stacked bar chart showing edible and poisonous proportions for each odor type. Only one category overlaps bewteen poisonous and edible mushrooms, whcich is odorless"
#| fig-pos: "H"
#| echo: false
#| message: false
#| warning: false

# Primary Author: Fatmah
# Reviewer: Heet
# PCIP: Plan to compare odor by class and create a plot.

library(tidyverse)
library(knitr)

mushroom_colors <- c(
  edible = "#2c7a4b",
  poisonous = "#8b3f8f"
)

mushroom_theme <- theme_bw()

odor_plot_data <- mushrooms_clean |>
  count(odor, classes, name = "count") |>
  group_by(odor) |>
  mutate(proportion = count / sum(count)) |>
  ungroup() |>
  arrange(-count)

ggplot(odor_plot_data, aes(y = odor, x = proportion, fill = classes)) +
  geom_col(position = "stack", color = "white", linewidth = 0.2) +
  scale_fill_manual(values = mushroom_colors) +
  mushroom_theme +
  labs(
    title = "Mushroom Edibility by Odor",
    subtitle = "Odor categories show strong differences in class composition.",
    x = "Proportion of samples",
    y = "Odor type",
    fill = "Class"
  )
```

While @tbl-odor displays the poisonous rate for each odor category so the visual pattern can be connected to exact summary values. We can see that Mushrooms with no odor is the only category that have both edible and poisonous. It is true that the majority of the odorless mushrooms are edible, but we believe that it is safer to use other characteristics to confirm instead of just relying on odor and smell.

```{r}
#| label: tbl-odor
#| tbl-cap: "Poisonous mushroom rate by odor category."
#| fig-alt: ""
#| echo: false
#| message: false
#| warning: false

# Primary Author: Fatmah
# Reviewer: Anchita
# PCIP: Plan to compute odor-specific poisonous rates and have a grouped summary.

library(tidyverse)
library(knitr)

odor_table <- mushrooms_clean |>
  count(odor, classes, name = "count") |>
  group_by(odor) |>
  mutate(rate = count / sum(count)) |>
  ungroup() |>
  filter(classes == "poisonous") |>
  arrange(-rate) |>
  mutate(`Poisonous Rate` = paste(round(rate * 100, 1), "%")) |>
  select(
    Odor = odor,
    `Poisonous Count` = count,
    `Poisonous Rate`
  )

odor_table |>
  kable()
```

### Gill Color:

Gill color in @fig-gill also shows noticeable differences between edible and poisonous mushrooms. Some colors appear more frequently in poisonous mushrooms, for example, Mushrooms with a buff gill color are always poisonous. While others are more common among edible mushrooms such as pink, brown, and white. These internal structural features can be used to further help distinguish mushroom classes.

```{r}
#| label: fig-gill
#| fig-cap: "Counts of edible and poisonous mushrooms by gill color."
#| fig-alt: "A grouped bar chart that compares edible and poisonous mushroom across differnt gill colors. Few gill colors are associated mostly with one class, while the rest are associated with both edible and poisonous mushrooms."
#| fig-pos: "H"
#| echo: false
#| message: false
#| warning: false

# Primary Author: Anchita
# Reviewer: Fatmah
# PCIP: Plan to compare gill colors and create a grouped bar chart to show the results.

library(tidyverse)
library(knitr)

gill_plot_data <- mushrooms_clean |>
  count(`gill-color`, classes, name = "count") |>
  arrange(-count)

ggplot(gill_plot_data, aes(x = `gill-color`, y = count, fill = classes)) +
  geom_col(position = "dodge", color = "white", linewidth = 0.2) +
  scale_fill_manual(values = mushroom_colors) +
  mushroom_theme +
  labs(
    title = "Edibility Patterns by Gill Color",
    subtitle = "Gill color categories differ in edible and poisonous counts.",
    x = "Gill color",
    y = "Number of mushrooms",
    fill = "Class"
  )
```

Looking at @tbl-gill, we can confirm that buff and green gill colors fall only under the poisonous mushroom class. In our previous case with odorless mushrooms, we can eliminate any odorless mushroom that have these colored gills.

```{r}
#| label: tbl-gill
#| tbl-cap: "Poisonous mushroom rate by gill color."
#| echo: false
#| message: false
#| warning: false

# Primary Author: Anchita
# Reviewer: Heet
# PCIP: Plan to summarize gill-color risk by calculating their poisonous rate.

gill_table <- mushrooms_clean |>
  count(`gill-color`, classes, name = "count") |>
  group_by(`gill-color`) |>
  mutate(rate = count / sum(count)) |>
  ungroup() |>
  filter(classes == "poisonous") |>
  arrange(-rate) |>
  mutate(`Poisonous Rate` = paste(round(rate * 100, 1), "%")) |>
  select(
    `Gill Color` = `gill-color`,
    `Poisonous Count` = count,
    `Poisonous Rate`
  )

gill_table |>
  kable()
```

### Bruising:

When we look at the bruising patterns between the classes of mushrooms in @fig-bruising, we notice that whether a mushroom bruises or not is not a good indicated to whether the mushroom is edible or not. that bruise and mushrooms that do not bruise show different class proportions, suggesting bruising may provide additional classification insight when combined with other traits.

```{r}
#| label: fig-bruising
#| fig-cap: "Mushroom edibility by bruising status."
#| fig-alt: "A stacked bar chart that compares edible and poisonous mushrooms for bruising and no bruising categories."
#| echo: false
#| message: false
#| warning: false

# Primary Author: Heet
# Reviewer: Fatmah
# PCIP: Plan to compare bruising status and see if can be used as another indicator of edible vs. poisonous mushroom.

bruising_plot_data <- mushrooms_clean |>
  count(`bruises?`, classes, name = "count") |>
  group_by(`bruises?`) |>
  mutate(proportion = count / sum(count)) |>
  ungroup()

ggplot(bruising_plot_data, aes(x = `bruises?`, y = proportion, fill = classes)) +
  geom_col(position = "stack", color = "white", linewidth = 0.2) +
  scale_fill_manual(values = mushroom_colors) +
  mushroom_theme +
  labs(
    title = "Mushroom Edibility by Bruising",
    subtitle = "Bruising status adds another categorical comparison.",
    x = "Bruising status",
    y = "Proportion of samples",
    fill = "Class"
  )
```

```{r}
#| label: tbl-bruising
#| tbl-cap: "Poisonous mushroom rate by bruising status."
#| echo: false
#| message: false
#| warning: false

# Primary Author: Heet
# Reviewer: Anchita
# PCIP: Plan to summarize bruising rates within the poisonous class.

bruise_table <- mushrooms_clean |>
  count(`bruises?`, classes, name = "count") |>
  group_by(`bruises?`) |>
  mutate(rate = count / sum(count)) |>
  ungroup() |>
  filter(classes == "poisonous") |>
  arrange(-rate) |>
  mutate(`Poisonous Rate` = paste(round(rate * 100, 1), "%")) |>
  select(
    Bruising = `bruises?`,
    `Poisonous Count` = count,
    `Poisonous Rate`
  )

bruise_table |>
  kable()
```

### Spore Print Color:

To further see how we can distinguish between edible and poisonous mushrooms, we decided to look at spore print colors. @fig-spore shows us that some spore print colors have a higher association with a specific mushroom class. Only 3 categories have only one of the classes. These categories can be used to eliminate if the mushroom we have is edible or poisonous.

```{r}
#| label: fig-spore
#| fig-cap: "Counts of edible and poisonous mushrooms by spore print color."
#| fig-alt: "A dot plot comparing mushroom counts by spore print color and class. Each point shows the number of mushrooms in a class for a spore print color category."
#| echo: false
#| message: false
#| warning: false

# Primary Author: Anchita
# Reviewer: Heet
# PCIP: Plan to compare the different spore print colors between each class.

spore_plot_data <- mushrooms_clean |>
  count(`spore-print-color`, classes, name = "count") |>
  arrange(-count)

ggplot(spore_plot_data, aes(x = count, y = `spore-print-color`, color = classes)) +
  geom_point(size = 5) +
  scale_color_manual(values = mushroom_colors) +
  mushroom_theme +
  labs(
    title = "Distribution by Spore Print Color",
    subtitle = "Dot positions show the count for each class within each category.",
    x = "Number of mushrooms",
    y = "Spore print color",
    color = "Class"
  )
```

Looking at @tbl-spore, we can see that green spores are the only true indicator that the mushroom is poisonous. all the other spore colors fail to fully indicate if the mushroom is edible or not becuase they overlap with poisonous mushrooms.

```{r}
#| label: tbl-spore
#| tbl-cap: "Poisonous mushroom rate by spore print color."
#| echo: false
#| message: false
#| warning: false

# Primary Author: Anchita
# Reviewer: Fatmah
# PCIP: Plan to summarize spore print color rates within the poisonous class.

spore_table <- mushrooms_clean |>
  count(`spore-print-color`, classes, name = "count") |>
  group_by(`spore-print-color`) |>
  mutate(rate = count / sum(count)) |>
  ungroup() |>
  filter(classes == "poisonous") |>
  arrange(-rate) |>
  mutate(`Poisonous Rate` = paste(round(rate * 100, 1), "%")) |>
  select(
    `Spore Print Color` = `spore-print-color`,
    `Poisonous Count` = count,
    `Poisonous Rate`
  )

spore_table |>
  kable()
```

## Limitations:

Although this dataset has many useful categorical features, it still has some limitations. It does not include geographic location, time period, weather, soil conditions, or detailed biological measurements that could provide more realistic context.

We also do not know how representative the 8,124 mushroom samples are. For example, it is unclear where the samples were collected, whether they came from many different locations, or whether they include a wide range of mushroom types.

Because of this, our exploratory analysis should be interpreted with caution. The results show patterns in the dataset, but they should not be treated as a safe field guide for mushroom identification or as a replacement for expert knowledge and real-world safety procedures.

## PCIP and Code Quality:

We used the PCIP process throughout the project: planning, coding, improving, and polishing.

In the planning stage, we focused on the mushroom class variable because it tells us whether each mushroom is edible or poisonous. We then chose categorical features that were useful, and easy to connect back to how to distinguish between the edible and poisonous mushrooms.

In the coding stage, we loaded the data file, renamed variables, decoded the category labels, and created summary tables. These summery table provided an easier understanding to the work we have done. Tasks were divided between us and carried separately between the team members.

In the improvement stage, we checked each other work and provided suggestions for each other to improve the final report. We also reused summary objects and revised our tables and plots so the results were easier to read.

In the polish stage, we followed each others comments and added captions, alt text, cross-references, and a code appendix. This made the report more organized, easier to follow, and easier to reproduce.

## Open Science and Reproducibility:

This project supports reproducibility by keeping the raw data, code, summary tables, and visualizations in one repository. The report can be easily rendered from the QMD file `Mushroom_project.qmd` saved in the repository.

We also used cross-referenced tables and figures so readers can connect the explanations we provided to actual data. We provided code chunks at the end of the report in the code appendix to show how the results were created.

## Conclusion:

In this project, we used exploratory data analysis to better understand how different mushroom features relate to whether a mushroom is edible or poisonous. The results showed that some features were much more useful than others. Odor was one of the strongest indicators, while gill color, bruising, and spore print color also helped separate the two classes.

One important takeaway is that one feature by itself is not always enough. Some categories overlap between edible and poisonous mushrooms, so it is safer and more useful to compare multiple features together. This helped us see which traits give stronger evidence when trying to distinguish between edible and poisonous mushrooms.

## Author Contributions"

**Fatmah:** Data import, cleaning, variables recording, documentation review.\
**Anchita:** Summary statistics, tables, visualizations, pattern analysis.\
**Heet:** Report writing, data provenance, FAIR/CARE discussion, conclusions, formatting and final review.

**NOTE:** All team members contributed to deciding the research focus, reviewing the dataset documentation, and interpreting the results.

## References

Lincoff, G. H. (1981). *The Audubon Society Field Guide to North American Mushrooms*. Alfred A. Knopf.

Schlimmer, J. (1987). *Mushroom Database*. UCI Machine Learning Repository. Dataset documentation included in this repository in `data/agaricus-lepiota.names`.

\newpage

## Code Appendix:

### Dataset cleaning code:

```{r}
#| label: appendix-load-clean-source
#| ref.label: load-and-clean-data
#| eval: false
```

### Plot creating code:

```{r}
#| label: appendix-odor-plot-source
#| ref.label: fig-odor
#| eval: false
```

```{r}
#| label: appendix-gill-plot-source
#| ref.label: fig-gill
#| eval: false
```

```{r}
#| label: appendix-bruising-plot-source
#| ref.label: fig-bruising
#| eval: false
```

```{r}
#| label: appendix-spore-plot-source
#| ref.label: fig-spore
#| eval: false
```

### Table creating code:

```{r}
#| label: appendix-class-source
#| ref.label: tbl-class
#| eval: false
```

```{r}
#| label: appendix-odor-table-source
#| ref.label: tbl-odor
#| eval: false
```

```{r}
#| label: appendix-gill-table-source
#| ref.label: tbl-gill
#| eval: false
```

```{r}
#| label: appendix-bruising-table-source
#| ref.label: tbl-bruising
#| eval: false
```

```{r}
#| label: appendix-spore-table-source
#| ref.label: tbl-spore
#| eval: false
```
